Appendix: Why the complement of AUC isn't a casewise CV criterion

### 2024-07-01 scratch/notes_georges/AUC/AUC_not_casewise-edited-1.Rmd
clean up scratch

---
documentclass: jss
author:
  - name: John Fox
    orcid: 0000-0002-1196-8012
    affiliation: 'McMaster University'
    address: |
      | Hamilton, Ontario, Canada
    email: \email{jfox@mcmaster.ca}
    url: https://www.john-fox.ca/
  - name: Georges Monette
    orcid: 0000-0000-0000-0000
    affiliation: 'York University'
    # To add another line, use \AND at the end of the previous one as above
    address: |
      | Toronto, Ontario, Canada
title:
  formatted: "\\pkg{cv}: An \\proglang{R} Package for Cross-Validation of Regression Models"
  # If you use tex in the formatted title, also supply version without
  plain:     "cv: An R Package for Cross-Validation of Regression Models"
  # For running headers, if needed
  short:     "\\pkg{cv}: Cross-Validation"
abstract: >
  The abstract of the article.
keywords:
  # at least one keyword must be supplied
  formatted: [cross-validation, regression analysis, model selection, "\\proglang{R}"]
  plain:     [cross-validation, regression analysis, model selection, R]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
bibliography: ["cv.bib"]
---

```{r, setup, include=FALSE}
options(prompt = 'R> ', continue = '+ ')
```

# Cross-validation

Cross-validation (CV) is an essentially simple and intuitively reasonable approach to estimating the predictive accuracy of regression models. CV is developed in many standard sources on regression modeling and "machine learning"---we particularly recommend @JamesEtAl:2021 [Secs. 5.1, 5.3]---and so we will describe the method only briefly here before taking up computational issues and some examples. See @ArlotCelisse:2010 for a wide-ranging, if technical, survey of cross-validation and related methods that emphasizes the statistical properties of CV.

Validating research by replication on independently collected data is a common scientific norm. Emulating this process in a single study by data-division is less common: The data are randomly divided into two, possibly equal-size, parts; the first part is used to develop and fit a statistical model; and then the second part is used to assess the adequacy of the model fit to the first part of the data. Data-division, however, suffers from two problems: (1) Dividing the data decreases the sample size and thus increases sampling error; and (2), even more disconcertingly, particularly in smaller samples, the results can vary substantially based on the random division of the data: See @Harrell:2015 [Sec. 5.3] for this and other remarks about data-division and cross-validation.

Cross-validation speaks to both of these issues. In CV, the data are randomly divided as equally as possible into several, say $k$, parts, called "folds." The statistical model is fit $k$ times, leaving each fold out in turn. Each fitted model is then used to predict the response variable for the cases in the omitted fold. A CV criterion or "cost" measure, such as the mean-squared error ("MSE") of prediction, is then computed using these predicted values. In the extreme $k = n$, the number of cases in the data, thus omitting individual cases and refitting the model $n$ times---a procedure termed "leave-one-out (LOO) cross-validation."

Because the $n$ models are each fit to $n - 1$ cases, LOO CV produces a nearly unbiased estimate of prediction error. The $n$ regression models are highly statistical dependent, however, based as they are on nearly the same data, and so the resulting estimate of prediction error has relatively large variance. In contrast, estimated prediction error for $k$-fold CV with $k = 5$ or $10$ (commonly employed choices) are somewhat biased but have smaller variance. It is also possible to correct $k$-fold CV for bias (see below).

# Examples

## Polynomial regression for the Auto data

The data for this example are drawn from the \pkg{ISLR2} package for \proglang{R}, associated with @JamesEtAl:2021. The presentation here is close (though not identical) to that in the original source [@JamesEtAl:2021 Secs. 5.1, 5.3], and it demonstrates the use of the `cv()` function in the \pkg{cv} package.[^boot]

[^boot]: @JamesEtAl:2021 use the `cv.glm()` function in the \pkg{boot} package [@CantyRipley2022; @DavisonHinkley:1997]. Despite its name, `cv.glm()` is an independent function and not a method of a `cv()` generic function.

The `Auto` dataset contains information about 392 cars:

```{r Auto}
data("Auto", package="ISLR2")
head(Auto)
dim(Auto)
```
With the exception of `origin` (which we don't use here), these variables are largely self-explanatory, except possibly for units of measurement: for details see `help("Auto", package="ISLR2")`.

We'll focus here on the relationship of `mpg` (miles per gallon) to `horsepower`, as displayed in the following scatterplot:

```{r mpg-horsepower-scatterplot}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "`mpg` vs `horsepower` for the `Auto` data"
plot(mpg ~ horsepower, data=Auto)
```
The relationship between the two variables is monotone, decreasing, and nonlinear. Following @JamesEtAl:2021, we'll consider approximating the relationship by a polynomial regression, with the degree of the polynomial $p$ ranging from 1 (a linear regression) to 10.[^log-trans] Polynomial fits  for $p = 1$ to $5$ are shown in the following figure:

[^log-trans]: Although it serves to illustrate the use of CV, a polynomial is probably not the best choice here. Consider, for example the scatterplot for log-transformed `mpg` and `horsepower`, produced by `plot(mpg ~ horsepower, data=Auto, log="xy")` (execution of which is left to the reader).

```{r mpg-horsepower-scatterplot-polynomials}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = "`mpg` vs `horsepower` for the `Auto` data"
plot(mpg ~ horsepower, data=Auto)
horsepower <- with(Auto, 
                   seq(min(horsepower), max(horsepower), 
                       length=1000))
for (p in 1:5){
  m <- lm(mpg ~ poly(horsepower,p), data=Auto)
  mpg <- predict(m, newdata=data.frame(horsepower=horsepower))
  lines(horsepower, mpg, col=p + 1, lty=p, lwd=2)
}
legend("topright", legend=1:5, col=2:6, lty=1:5, lwd=2,
       title="Degree", inset=0.02)
```
The linear fit is clearly inappropriate; the fits for $p = 2$ (quadratic) through $4$ are very similar; and the fit for $p = 5$ may over-fit the data by chasing one or two relatively high `mpg` values at the right (but see the CV results reported below).

The following graph shows two measures of estimated (squared) error as a function of polynomial-regression degree: The mean-squared error ("MSE"), defined as $\mathsf{MSE} = \frac{1}{n}\sum_{i=1}^n (y_i - \widehat{y}_i)^2$, and the usual residual variance, defined as $\widehat{\sigma}^2 = \frac{1}{n - p - 1} \sum_{i=1}^n (y_i - \widehat{y}_i)^2$. The former necessarily declines with $p$ (or, more strictly, can't increase with $p$), while the latter gets slightly larger for the largest values of $p$, with the "best" value, by a small margin, for $p = 7$.

```{r mpg-horsepower-MSE-se}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = "Estimated squared error as a function of polynomial degree, $p$"
library("cv") # for mse() and other functions

var <- mse <- numeric(10)
for (p in 1:10){
  m <- lm(mpg ~ poly(horsepower, p), data=Auto)
  mse[p] <- mse(Auto$mpg, fitted(m))
  var[p] <- summary(m)$sigma^2
}

plot(c(1, 10), range(mse, var), type="n",
     xlab="Degree of polynomial, p",
     ylab="Estimated Squared Error")
lines(1:10, mse, lwd=2, lty=1, col=2, pch=16, type="b")
lines(1:10, var, lwd=2, lty=2, col=3, pch=17, type="b")
legend("topright", inset=0.02,
       legend=c(expression(hat(sigma)^2), "MSE"),
       lwd=2, lty=2:1, col=3:2, pch=17:16)
```
The code for this graph uses the `mse()` function from the \pkg{cv} package to compute the MSE for each fit.

### Using cv()

The generic `cv()` function has an `"lm"` method, which by default performs $k = 10$-fold CV:
```{r cv-lm-1}
m.auto <- lm(mpg ~ poly(horsepower, 2), data=Auto)
summary(m.auto)

cv(m.auto)
```
The \code{"lm"} method by default uses \code{mse()} as the CV criterion and the Woodbury matrix identity to update the regression with each fold deleted without having literally to refit the model. Computational details are discussed in the final section of this vignette. The function reports the CV estimate of MSE, a biased-adjusted estimate of the MSE (the bias adjustment is explained in the final section), and the MSE is also computed for the original, full-sample regression. Because the division of the data into 10 folds is random, \code{cv()} explicitly (randomly) generates and saves a seed for \proglang{R}'s pseudo-random number generator, to make the results replicable. The user can also specify the seed directly via the \code{seed} argument to \code{cv()}.

To perform LOO CV, we can set the \code{k} argument to \code{cv()} to the number of cases in the data, here \code{k=392}, or, more conveniently, to \code{k="loo"} or \code{k="n"}:

```{r cv.lm-2`}
cv(m.auto, k="loo")
```
For LOO CV of a linear model, \code{cv()} by default uses the hatvalues from the model fit to the full data for the LOO updates, and reports only the CV estimate of MSE. Alternative methods are to use the Woodbury matrix identity or the "naive" approach of literally refitting the model with each case omitted. All three methods produce exact results for a linear model (within the precision of floating-point computations):
```{r cv.lm-3}
cv(m.auto, k="loo", method="naive")

cv(m.auto, k="loo", method="Woodbury")
```
The \code{"naive"} and \code{"Woodbury"} methods also return the bias-adjusted estimate of MSE and the full-sample MSE, but bias isn't an issue for LOO CV.

This is a small regression problem and all three computational approaches are essentially instantaneous, but it is still of interest to investigate their relative speed. In this comparison, we include the \code{cv.glm()} function from the \pkg{boot} package, which takes the naive approach, and for which we have to fit the linear model as an equivalent Gaussian GLM. We use the \code{microbenchmark()} function from the package of the same name for the timings [@Mersmann:2023]:
```{r cv.lm.timings, cache=TRUE}
m.auto.glm <- glm(mpg ~ poly(horsepower, 2), data=Auto)
boot::cv.glm(Auto, m.auto.glm)$delta

microbenchmark::microbenchmark(
  hatvalues = cv(m.auto, k="loo"),
  Woodbury = cv(m.auto, k="loo", method="Woodbury"),
  naive = cv(m.auto, k="loo", method="naive"),
  cv.glm = boot::cv.glm(Auto, m.auto.glm),
  times=10
)
```
On our computer, using the hatvalues is about an order of magnitude faster than employing Woodbury matrix updates, and more than two orders of magnitude faster than refitting the model.[^microbenchmark]

[^microbenchmark]: Out of impatience, we asked \code{microbenchmark()} to execute each command only 10 times rather than the default 100. With the exception of the last columns, the output is self-explanatory. The last column shows which methods have average timings that are statistically distinguishable. Because of the small number of repetitions (i.e., 10), the \code{"hatvalues"} and \code{"Woodbury"} methods aren't distinguishable, but the difference between these methods persists when we perform more repetitions---we invite the reader to redo this computation with the default \code{times=100} repetitions.

### Comparing competing models

The \code{cv()} function also has a method that can be applied to a list of regression models for the same data, composed using the \code{models()} function. For $k$-fold CV, the same folds are used for the competing models, which reduces random error in their comparison. This result can also be obtained by specifying a common seed for \proglang{R}'s random-number generator while applying \code{cv()} separately to each model, but employing a list of models is more convenient for both $k$-fold and LOO CV (where there is no random component to the composition of the $n$ folds).

We illustrate with the polynomial regression models of varying degree for the \code{Auto} data (discussed previously), beginning by fitting and saving the 10 models:
```{r polyomial-models}
for (p in 1:10){
  assign(paste0("m.", p),
         lm(mpg ~ poly(horsepower, p), data=Auto))
}
objects(pattern="m\\.[0-9]")
summary(m.2) # for example, the quadratic fit
```
We then apply \code{cv()} to the list of 10 models (the \code{data} argument is required):
```{r polynomial-regression-CV}
# 10-fold CV
cv.auto.10 <- cv(models(m.1, m.2, m.3, m.4, m.5,
                     m.6, m.7, m.8, m.9, m.10),
              data=Auto, seed=2120)
cv.auto.10[1:2] # for the linear and quadratic models

# LOO CV
cv.auto.loo <- cv(models(m.1, m.2, m.3, m.4, m.5,
                        m.6, m.7, m.8, m.9, m.10),
                 data=Auto, k="loo")
cv.auto.loo[1:2] # linear and quadratic models
```
Because we didn't supply names for the models in the calls to the \code{models()} function, the names \code{model.1}, \code{model.2}, etc., are generated by the function.

Finally, we extract and graph the adjusted MSEs for $10$-fold CV and the MSEs for LOO CV:
```{r polynomial-regression-CV-graph}
#| out.width = "100%",
#| fig.height = 5,
#| fig.cap = "Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$"
cv.mse.10 <- sapply(cv.auto.10, function(x) x[["adj CV crit"]])
cv.mse.loo <- sapply(cv.auto.loo, function(x) x[["CV crit"]])
plot(c(1, 10), range(cv.mse.10, cv.mse.loo), type="n",
     xlab="Degree of polynomial, p",
     ylab="Cross-Validated MSE")
lines(1:10, cv.mse.10, lwd=2, lty=1, col=2, pch=16, type="b")
lines(1:10, cv.mse.loo, lwd=2, lty=2, col=3, pch=17, type="b")
legend("topright", inset=0.02,
       legend=c("10-Fold CV", "LOO CV"),
       lwd=2, lty=2:1, col=3:2, pch=17:16)
```

Alternatively, we can use the \code{plot()} method for \code{"cvModList"} objects to compare the models, though with separate graphs for 10-fold and LOO CV:

```{r, polynomial-regression-CV-graph-2, fig.show="hold"}
#| out.width = "45%",
#| fig.height = 6,
#| fig.cap = "Cross-validated 10-fold and LOO MSE as a function of polynomial degree, $p$"
plot(cv.auto.10, main="Polynomial Regressions, 10-Fold CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
plot(cv.auto.loo, main="Polynomial Regressions, LOO CV",
     axis.args=list(labels=1:10), xlab="Degree of Polynomial, p")
```

In this example, 10-fold and LOO CV produce generally similar results, and also results that are similar to those produced by the estimated error variance $\widehat{\sigma}^2$ for each model, reported above (except for the highest-degree polynomials, where the CV results more clearly suggest over-fitting).

## Logistic regression for the Mroz data

The \code{Mroz} data set from the \pkg{carData} package [associated with @FoxWeisberg:2019] has been used by several authors to illustrate binary logistic regression; see, in particular @FoxWeisberg:2019. The data were originally drawn from the U.S. Panel Study of Income Dynamics and pertain to married women. Here are a few cases in the data set:
```{r Mroz-data}
data("Mroz", package="carData")
head(Mroz, 3)
tail(Mroz, 3)
```
The response variable in the logistic regression is \code{lfp}, labor-force participation, a factor coded \code{"yes"} or \code{"no"}. The remaining variables are predictors:

* \code{k5}, number of children 5 years old of younger in the woman's household;
* \code{k618}, number of children between 6 and 18 years old;
* \code{age}, in years;
* \code{wc}, wife's college attendance, \code{"yes"} or \code{"no"};
* \code{hc}, husband's college attendance;
* \code{lwg}, the woman's log wage rate if she is employed, or her *imputed* wage rate, if she is not [a variable that @FoxWeisberg:2019 show is problematically defined]; and
* \code{inc}, family income, in $1000s, exclusive of wife's income.

We use the \code{glm()} function to fit a binary logistic regression to the \code{Mroz} data:
```{r Mroz-logistic-regresion}
m.mroz <- glm(lfp ~ ., data=Mroz, family=binomial)
summary(m.mroz)

BayesRule(ifelse(Mroz$lfp == "yes", 1, 0), 
          fitted(m.mroz, type="response"))
```
In addition to the usually summary output for a GLM, we show the result of applying the \code{BayesRule()} function from the \pkg{cv} package to predictions derived from the fitted model. Bayes rule, which predicts a "success" in a binary regression model when the fitted probability of success [i.e., $\phi = \Pr(y = 1)$] is $\widehat{\phi} \ge .5$ and a "failure" if $\widehat{\phi} < .5$.[^BayesRule] The first argument to \code{BayesRule()} is the binary {0, 1} response, and the second argument is the predicted probability of success. \code{BayesRule()} returns the proportion of predictions that are *in error*, as appropriate for a "cost" function.

[^BayesRule]: \code{BayesRule()} does some error checking; \code{BayesRule2()} is similar, but omits the error checking, and so can be faster for large problems.

The value returned by \code{BayesRule()} is associated with an "attribute" named \code{"casewise loss"} and set to \code{"y != round(yhat)"}, signifying that the Bayes rule CV criterion is computed as the mean of casewise values, here 0 if the prediction for a case matches the observed value and 1 if it does not (signifying a prediction error). The \code{mse()} function for numeric responses is also calculated as a casewise average. Some other criteria, such as the median absolute error, computed by the \code{medAbsErr()} function in the \pkg{cv} package, aren't averages of casewise components. The distinction is important because, to our knowledge, the statistical theory of cross-validation, for example, in @DavisonHinkley:1997, @BatesHastieTibshirani:2023, and @ArlotCelisse:2010, is developed for CV criteria like MSE that are means of casewise components. As a consequence, we limit computation of bias adjustment and confidence intervals (see below) to criteria that are casewise averages.

In this example, the fitted logistic regression incorrectly predicts 31% of the responses; we expect this estimate to be optimistic given that the model is used to "predict" the data to which it is fit.

The \code{"glm"} method for \code{cv()} is largely similar to the \code{"lm"} method, although the default algorithm, selected explicitly by \code{method="exact"}, refits the model with each fold removed (and is thus equivalent to \code{method="naive"} for \code{"lm"} models). For generalized linear models, \code{method="Woodbury"} or (for LOO CV) \code{method="hatvalues"} provide approximate results (see the last section of the vignette for details):

```{r cv-Mroz-10-fold}
cv(m.mroz, criterion=BayesRule, seed=248)

cv(m.mroz, criterion=BayesRule, seed=248, method="Woodbury")
```
To ensure that the two methods use the same 10 folds, we specify the seed for \proglang{R}'s random-number generator explicitly; here, and as is common in our experience, the \code{"exact"} and \code{"Woodbury"} algorithms produce nearly identical results. The CV estimates of prediction error are slightly higher than the estimate based on all of the cases.

The printed output includes a 95% confidence interval for the bias-adjusted Bayes rule CV criterion. @BatesHastieTibshirani:2023 show that these confidence intervals are unreliable for models fit to small samples, and by default \code{cv()} computes them only when the sample size is 400 or larger and when the CV criterion employed is an average of casewise components, as is the case for Bayes rule. See the final section of the vignette for details of the computation of confidence intervals for bias-adjusted CV criteria.

Here are results of applying LOO CV to the Mroz model, using both the exact and the approximate methods:
```{r cv-Mroz-LOO}
cv(m.mroz, k="loo", criterion=BayesRule)

cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury")

cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues")
```
To the number of decimal digits shown, the three methods produce identical results for this example.

As for linear models, we report some timings for the various \code{cv()` methods of computation in LOO CV as well as for the \code{cv.glm()} function from the \pkg{boot} package (which, recall, refits the model with each case removed, and thus is comparable to \code{cv()} with \code{method="exact"}):
```{r glm.timings, cache=TRUE}
microbenchmark::microbenchmark(
  hatvalues=cv(m.mroz, k="loo", criterion=BayesRule, method="hatvalues"),
  Woodbury=cv(m.mroz, k="loo", criterion=BayesRule, method="Woodbury"),
  exact=cv(m.mroz, k="loo", criterion=BayesRule),
  cv.glm=boot::cv.glm(Mroz, m.mroz,
               cost=BayesRule),
  times=10)
```
There is a substantial time penalty associated with exact computations.

# Cross-validating mixed-effects models

The fundamental analogy for cross-validation is to the collection of new data. That is, predicting the response in each fold from the model fit to data in the other folds is like using the model fit to all of the data to predict the response for new cases from the values of the predictors for those new cases. As we explained, the application of this idea to independently sampled cases is straightforward---simply partition the data into random folds of equal size and leave each fold out in turn, or, in the case of LOO CV, simply omit each case in turn.

In contrast, mixed-effects models are fit to *dependent* data, in which cases as clustered, such as hierarchical data, where the clusters comprise higher-level units (e.g., students clustered in schools), or longitudinal data, where the clusters are individuals and the cases repeated observations on the individuals over time.[^crossed-effects]

[^crossed-effects]: There are, however, more complex situations that give rise to so-called *crossed* (rather than *nested*) random effects. For example, consider students within classes within schools. In primary schools, students typically are in a single class, and so classes are nested within schools. In secondary schools, however, students typically take several classes and students who are together in a particular class may not be together in other classes; consequently, random effects based on classes within schools are crossed. The \code{lmer()} function in the \pkg{lme4} package is capable of modeling both nested and crossed random effects, and the \code{cv()} methods for mixed models in the \pkg{cv} package pertain to both nested and crossed random effects. We present an example of the latter later in the vignette.

We can think of two approaches to applying cross-validation to clustered data:[^cv-faq]

[^cv-faq]: We subsequently discovered that @Vehtari:2023 [Section 8] makes similar points.

1. Treat CV as analogous to predicting the response for one or more cases in a *newly observed cluster*. In this instance, the folds comprise one or more whole clusters; we refit the model with all of the cases in clusters in the current fold removed; and then we predict the response for the cases in clusters in the current fold. These predictions are based only on fixed effects because the random effects for the omitted clusters are presumably unknown, as they would be for data on cases in newly observed clusters.

2. Treat CV as analogous to predicting the response for a newly observed case in an *existing cluster*. In this instance, the folds comprise one or more individual cases, and the predictions can use both the fixed and random effects.

## Example: The High-School and Beyond data

Following their use by @RaudenbushBryk:2002, data from the 1982 *High School and Beyond* (HSB) survey have become a staple of the literature on mixed-effects models. The HSB data are used by @FoxWeisberg:2019 [Sec. 7.2.2] to illustrate the application of linear mixed models to hierarchical data, and we'll closely follow their example here.

The HSB data are included in the \code{MathAchieve} and \code{MathAchSchool} data sets in the \pkg{nlme} package  [@PinheiroBates:2000]. \code{MathAchieve} includes individual-level data on 7185  students in 160 high schools, and \code{MathAchSchool} includes school-level data:
```{r HSB-data}
data("MathAchieve", package="nlme")
dim(MathAchieve)
head(MathAchieve, 3)
tail(MathAchieve, 3)

data("MathAchSchool", package="nlme")
dim(MathAchSchool)
head(MathAchSchool, 2)
tail(MathAchSchool, 2)
```
The first few students are in school number 1224 and the last few in school 9586.

We'll use only the \code{School}, \code{SES} (students' socioeconomic status), and \code{MathAch} (their score on a standardized math-achievement test) variables in the \code{MathAchieve} data set, and \code{Sector} (\code{"Catholic"} or \code{"Public"}) in the \code{MathAchSchool} data set.

Some data-management is required before fitting a mixed-effects model to the HSB data, for which we use the \pkg{dplyr} package [@WickhamEtAl:2023]:
```{r HSB-data-management, cache=TRUE}
library("dplyr")
MathAchieve %>% group_by(School) %>%
  summarize(mean.ses = mean(SES)) -> Temp
Temp <- merge(MathAchSchool, Temp, by="School")
HSB <- merge(Temp[, c("School", "Sector", "mean.ses")],
             MathAchieve[, c("School", "SES", "MathAch")], by="School")
names(HSB) <- tolower(names(HSB))

HSB$cses <- with(HSB, ses - mean.ses)
```
In the process, we created two new school-level variables: \code{meanses}, which is the average SES for students in each school; and \code{cses}, which is school-average SES centered at its mean. For details, see @FoxWeisberg:2019 [Sec. 7.2.2].

Still following Fox and Weisberg, we proceed to use the \code{lmer()} function in the \pkg{lme4} package [@BatesEtAl:2015] to fit a mixed model for math achievement to the HSB data:
```{r HSB-lmer, cache=TRUE}
library("lme4")
hsb.lmer <- lmer(mathach ~ mean.ses*cses + sector*cses
                   + (cses | school), data=HSB)
summary(hsb.lmer, correlation=FALSE)
```

We can then cross-validate at the cluster (i.e., school) level,
```{r HSB-lmer-CV-cluster, cache=TRUE}
cv(hsb.lmer, k=10, clusterVariables="school", seed=5240)
```
or at the case (i.e., student) level,
```{r HSB-lmer-CV-case, cache=TRUE}
cv(hsb.lmer, seed=1575)
```
For cluster-level CV, the \code{clusterVariables} argument tells \code{cv()} how the clusters are defined. Were there more than one clustering variable, say classes within schools, these would be provided as a character vector of variable names: \code{clusterVariables = c("school", "class")}. For cluster-level CV, the default is \code{k = "loo"}, that is, leave one cluster out at a time; we instead specify \code{k = 10} folds of clusters, each fold therefore comprising $160/10 = 16$ schools.

If the \code{clusterVariables} argument is omitted, then case-level CV is employed, with \code{k = 10} folds as the default, here each with $7185/10 \approx 719$ students. Notice that one of the 10 models refit with a fold removed failed to converge. Convergence problems are common in mixed-effects modeling. The apparent issue here is that an estimated variance component is close to or equal to 0, which is at a boundary of the parameter space. That shouldn't disqualify the fitted model for the kind of prediction required for cross-validation.

There is also a \code{cv()} method for linear mixed models fit by the \code{lme()} function in the \pkg{nlme} package, and the arguments for \code{cv()} in this case are the same as for a model fit by \code{lmer()} or \code{glmer()}. We illustrate with the mixed model fit to the HSB data:
```{r hsb-lme, cache=TRUE}
library(nlme)
hsb.lme <- lme(mathach ~ mean.ses*cses + sector*cses,
                 random = ~ cses | school, data=HSB,
               control=list(opt="optim"))
summary(hsb.lme)

cv(hsb.lme, k=10, clusterVariables="school", seed=5240)

cv(hsb.lme, seed=1575)
```
We used the same random-number generator seeds as in the previous example cross-validating the model fit by \code{lmer()}, and so the same folds are employed in both cases.[^optimizer] The estimated covariance components and fixed effects in the summary output differ slightly between the \code{lmer()} and \code{lme()} solutions, although both functions seek to maximize the REML criterion. This is, of course, to be expected when different algorithms are used for numerical optimization.  To the precision reported, the cluster-level CV results for the \code{lmer()} and \code{lme()} models are identical, while the case-level CV results are very similar but not identical.

[^optimizer]: The observant reader will notice that we set the argument \code{control=list(opt="optim")} in the call to \code{lme()}, changing the optimizer employed from the default \code{"nlminb"}. We did this because with the default optimizer, \code{lme()} encountered the same convergence issue as \code{lmer()}, but rather than issuing a warning, \code{lme()} failed, reporting an error. As it turns out, setting the optimizer to \code{"optim"} avoids this problem.

## Example: Contrived hierarchical data

We introduce an artificial data set that exemplifies aspects of cross-validation particular to hierarchical models. Using this data set, we show that model comparisons employing cluster-based and those employing case-based cross-validation may not agree on a "best" model. Furthermore, commonly used measures of fit, such as mean-squared error, do not necessarily become smaller as models become larger, even when the models are nested, and even when the measure of fit is computed for the whole data set.

Consider a researcher studying improvement in a skill, yodeling, for example, among students enrolled in a four-year yodeling program. The plan is to measure each student's skill level at the beginning of the program and every year thereafter until the end of the program, resulting in five annual measurements for each student. It turns out that yodeling appeals to students of all ages, and students enrolling in the program range in age from 20 to 70. Moreover, participants' untrained yodeling skill is similar at all ages, as is their rate of progress with training. All students complete the four-year program.

The researcher, who has more expertise in yodeling than in modeling, decides to model the response, $y$, yodeling skill, as a function of age, $x$, reasoning that students get older during their stay in the program, and (incorrectly) that age can serve as a proxy for elapsed time. The researcher knows that a mixed model should be used to account for clustering due to the expected similarity of measurements taken from each student.

We start by generating the data, using parameters consistent with the description above and meant to highlight the issues that arise in cross-validating mixed-effects models:[^1]

[^1]: We invite the interested reader to experiment with varying the parameters of our example.

```{r data}
# Parameters:
set.seed(9693) 
Nb <- 100     # number of groups
Nw <- 5       # number of individuals within groups
Bb <- 0       # between-group regression coefficient on group mean
SDre <- 2.0   # between-group SD of random level relative to group mean of x
SDwithin <- 0.5  # within group SD
Bw <- 1          # within group effect of x
Ay <- 10         # intercept for response
Ax <- 20         # starting level of x
Nx <- Nw*10      # number of distinct x values

Data <- data.frame(
  group = factor(rep(1:Nb, each=Nw)),
  x = Ax + rep(1:Nx, length.out = Nw*Nb)
) |>
  within(
    {
      xm  <- ave(x, group, FUN = mean) # within-group mean
      y <- Ay +
        Bb * xm +                    # contextual effect
        Bw * (x - xm) +              # within-group effect
        rnorm(Nb, sd=SDre)[group] +  # random level by group
        rnorm(Nb*Nw, sd=SDwithin)    # random error within groups
    }
  )
```

Here is a scatterplot of the data for a representative group of 10 (without loss of generality, the first 10) of 100 students, showing the 95% concentration ellipse for each cluster:[^2]

[^2]: We find it convenient to use the \pkg{lattice} [@Sarkar:2008] and \pkg{latticeExtra} [@SarkarAndrews:2022] packages for this and other graphs in this section.

```{r plot1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Hierarchical data set, showing the first 10 of 100 students."
library("lattice")
library("latticeExtra")
plot <- xyplot(y ~ x, data=Data[1:Nx, ], group=group,
               ylim=c(4, 16),
               par.settings=list(superpose.symbol=list(pch=1, cex=0.7))) +
    layer(panel.ellipse(..., center.cex=0))
plot # display graph
```

The between-student effect of age is 0 but the within-student effect is 1. Due to the large variation in ages between students, the least-squares regression of yodeling skill on age (for the 500 observations among all 100 students) produces an estimated slope close to 0 (though with a small $p$-value), because the slope is heavily weighted toward the between-student effect:

```{r}
summary(lm(y ~ x, data=Data))
```

The initial mixed-effects model that we fit to the data is a simple random-intercepts model:
```{r include=FALSE, echo=FALSE}
library(lme4) # necessary for some reason to knit vignette in RStudio, harmless otherwise
```
```{r}
# random intercept only:
mod.0 <- lmer(y ~ 1 + (1 | group), Data)
summary(mod.0)
```

We will shortly consider three other, more complex, mixed models; because of data-management considerations, it is convenient to fit them now, but we defer discussion of these models:

```{r}
# effect of x and random intercept:
mod.1 <- lmer(y ~ x + (1 | group), Data)

# effect of x, contextual (student) mean of x, and random intercept:
mod.2 <- lmer(y ~ x + xm + (1 | group), Data)
        # equivalent to y ~ I(x - xm) + xm + (1 | group)

# model generating the data (where Bb = 0)
mod.3 <- lmer(y ~ I(x - xm) + (1 | group), Data)
```

We proceed to obtain predictions from the random-intercept model (\code{mod.0}) and the other models (\code{mod.1}, \code{mod.2}, and \code{mod.3}) based on fixed effects alone, as would be used for cross-validation based on clusters (i.e., students), and for fixed and random effects---so-called best linear unbiased predictions or BLUPs---as would be used for cross-validation based on cases (i.e., occasions within students):

```{r}
Data <- within(Data, {
  fit_mod0.fe <- predict(mod.0, re.form = ~ 0) # fixed effects only
  fit_mod0.re <- predict(mod.0) # fixed and random effects (BLUPs)
  fit_mod1.fe <- predict(mod.1, re.form = ~ 0)
  fit_mod1.re <- predict(mod.1)
  fit_mod2.fe <- predict(mod.2, re.form = ~ 0)
  fit_mod2.re <- predict(mod.2)
  fit_mod3.fe <- predict(mod.3, re.form = ~ 0)
  fit_mod3.re <- predict(mod.3)
})
```

We then prepare the data for plotting:

```{r}
Data_long <- reshape(Data[1:Nx, ], direction = "long", sep = ".", 
              timevar = "effect", varying = grep("\\.", names(Data[1:Nx, ])))
Data_long$id <- 1:nrow(Data_long)
Data_long <- reshape(Data_long, direction = "long", sep = "_", 
              timevar = "modelcode",  varying = grep("_", names(Data_long)))
Data_long$model <- factor(
  c("~ 1", "~ 1 + x", "~ 1 + x + xm", "~ 1 + I(x - xm)")
  [match(Data_long$modelcode, c("mod0", "mod1", "mod2", "mod3"))]
)
```

Predictions based on the random-intercept model \code{mod.0} for the first 10 students are shown in the following graph:

```{r plot-fits-mod0}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the random intercept model."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod0" &  effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + (1 | group)",
  key=list(
    corner=c(0.05, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

The fixed-effect predictions for the various individuals are identical---the estimated fixed-effects intercept or estimated general mean of $y$---while the BLUPs are the sums of the fixed-effects intercept and the random intercepts, and are only slightly shrunken towards the general mean. Because in our artificial data there is no population relationship between age and skill, the fixed-effect-only predictions and the BLUPs are not very different.

Our next model, \code{mod.1}, includes a fixed intercept and fixed effect of \code{x} along with a random intercept:

```{r}
summary(mod.1)
```

Predictions from this model appear in the following graph:

```{r plot-fits-mod1}
#| out.width = "100%",
#| fig.height = 6,
#| fig.cap = "Predictions from the model with random intercepts and $x$ as a fixed-effect predictor."
(plot +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "fe"),
         groups=group, type="l", lwd=2) +
  xyplot(fit ~ x, subset(Data_long, modelcode == "mod1" & effect == "re"),
         groups=group, type="l", lwd=2, lty=3)
) |> update(
  main="Model: y ~ 1 + x + (1 | group)",
  ylim=c(-15, 35),
  key=list(
    corner=c(0.95, 0.05),
    text=list(c("fixed effects only","fixed and random")),
    lines=list(lty=c(1, 3))))
```

The BLUPs fit the observed data very closely, but predictions based on the fixed effects alone, with a common intercept and slope for all clusters, are very poor---indeed, much worse than the fixed-effects-only predictions based on the simpler random-intercept model, \code{mod.0}. We therefore anticipate (and show later in this section) that case-based cross-validation will prefer \code{mod1} to \code{mod0}, but that cluster-based cross-validation will prefer \code{mod0} to \code{mod1}.

Our third model, \code{mod.2}, includes the contextual effect of $x$---that is, the cluster mean \code{xm]---along with $x$ and the intercept in the fixed-effect part of the model, and a random intercept:

```{r}
summary(mod.2)
```

This model is equivalent to fitting \code{y ~ I(x - xm) + xm + (1 | group)}, which is the model that generated the data once the coefficient of the contextual predictor \code{xm} is set to 0 (as it is in \code{mod.3], discussed below).

>>>>

---
title: "Appendix: Why the complement of AUC isn't a casewise CV criterion"
output: html_document
---
